A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-13 code/supplements.Rmd

### Supplementary Figure S1
```{r supplementary-figure-s1, fig.width = 9, fig.height= 8, fig.align='center'}
#out.width = '90%',
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data

@@

ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = project_code)) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 23)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
```
\newpage
### Supplementary Figure S2
```{r supplementary-figure-s2-data-generation, fig.width = 9, fig.height= 8, fig.align = "center"}
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = Subtype_Selected)) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
```
\newpage
### Supplementary Figure S3
```{r supplementary-figure-s3-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

@@

ggplot(selected_mut, aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 20)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ factor(variable, levels = c("MSH2", "MSH6", "MLH1", "MLH3", "PMS1", "PMS2", "BRIP1", "RAD51", "CHEK2", "APC", "TET2", "TET3", "TERT", "TEP1", "DKC1")), ncol = 5) ## facet by gene
```
\newpage
### Supplementary Figure S4
```{r supplementary-figure-s4-data-generation, eval = FALSE}
load("../data/gdc_reactome_path_analysis.Rda") ## load data
umap_cancertype <- fread("../data/umap_3d_coors.tsv") 
n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway

@@

```
\newpage
### Supplementary Figure S5
```{r, supplemental-s5-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

@@

umap_clust178 + pik3ca + kras + tp53 + plot_layout(design = layout)
```
\newpage
### Supplementary Table 1
```{r supplemental-table-st1-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gene_data.Rda")
load("../data/gdc_gene_exp_adj.Rda")

>>>>

\newpage
### Supplementary Figure S1
```{r supplementary-figure-s1, fig.width = 9, fig.height= 8, fig.align='center'}
#out.width = '90%',
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data

@@

ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = project_code)) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 23)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer_type
```
\pagebreak
### Supplementary Figure S2
```{r supplementary-figure-s2-data-generation, fig.width = 9, fig.height= 8, fig.align = "center"}
umap_cancertype <- fread("../data/umap_3d_coors.tsv") ## load data
umap_bg <- umap_cancertype[,-"project_code"] ## get background data for facet

ggplot(umap_cancertype, aes(x = plot_y, y = plot_x, colour = Subtype_Selected)) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ project_code, ncol = 5) ## facet by cancer type
```
\pagebreak
### Supplementary Figure S3
```{r supplementary-figure-s3-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

@@

ggplot(selected_mut, aes(x = plot_y, y = plot_x, colour = as.character(value))) + 
  geom_point(data = umap_bg, aes(x = plot_y, y = plot_x), colour = "grey", size = 0.25) +
  geom_point(size = 0.25) +
  theme_classic() +
  coord_fixed() +
  xlab("UMAP 1") + ylab("UMAP 2") +
  scale_colour_manual(values = rep("firebrick3", 20)) +
  theme(axis.text.x = element_blank(),
    axis.text.y = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    axis.line = element_blank(),
    legend.position = "none") +
  facet_wrap(~ factor(variable, levels = c("MSH2", "MSH6", "MLH1", "MLH3", "PMS1", "PMS2", "BRIP1", "RAD51", "CHEK2", "APC", "TET2", "TET3", "TERT", "TEP1", "DKC1")), ncol = 5) ## facet by gene
```
\pagebreak
### Supplementary Figure S4
```{r supplementary-figure-s4-data-generation, eval = FALSE}
load("../data/gdc_reactome_path_analysis.Rda") ## load data
umap_cancertype <- fread("../data/umap_3d_coors.tsv") 
n_path <- 377 ## how many pathways do we want to look at starting from the most to least mutated pathway

@@

```
\pagebreak
### Supplementary Figure S5
```{r, supplemental-s5-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gene_data.Rda")

@@

umap_clust178 + pik3ca + kras + tp53 + plot_layout(design = layout)
```
\pagebreak
### Supplementary Table 1
```{r supplemental-table-st1-data-generation, eval = FALSE}
load("../data/gdc_tumor_coding_complete.Rdata")
load("../data/gdc_reactome_path_analysis.Rda")
load("../data/gene2_level2_lookup.Rda")
load("../data/pathname2level2.Rda")
load("../data/gene_data.Rda")
load("../data/gdc_gene_exp_adj.Rda")

### 2020-08-13 text/manuscript.Rmd

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi=600, warning = FALSE)
suppressPackageStartupMessages(c(
  library(rstan),
  library(foreach),
  library(sqldf),
  library(tidyverse),
  library(ggthemes),
  library(data.table),
  library(biomaRt),
  library(dplyr),
  library(GenomicRanges),
  library(SummarizedExperiment),
  library(reactome.db),
  library(plyr),
  library(cluster),
  library(pheatmap),
  library(clusterProfiler),
  library(org.Hs.eg.db),
  library(dbscan),
  library(class),
  library(TCGAbiolinks),
  library(patchwork),
  library(ggplotify),
  library(ComplexHeatmap),
  library(knitr),
  library(kableExtra),
  library(UpSetR),
  library(cowplot),
  library(doRNG),
  library(uwot),
  library(janitor),
  library(FactoMineR),
  library(seriation),
  library(rtracklayer),
  library(poibin),
  library(flextable)
))
```

@@

<strike>
### Gene-Pathway redundancy contributes to cluster identity

Since pathways share some common driver genes, we investigated whether
tumors in each class  have multiple mutations across signaling pathways, or
whether a small number of driver genes account for enrichment. The split between
clusters 7 and 8 is largely driven by mutations in _PI3K_ and its
homologs and _Ras_ genes, respectively (compare _PIK3CA_ and _KRAS_ panels of
**supplementary figure S5**). Furthermore, cluster 1 is distinguished
from clusters 2-10 by general lack of P53 mutations (compare **Figure 1f** with the
_TP53_ panel of **Supplementary Figure S5**).

Cluster 2 exhibited the highest pathway enrichment scores of the ten major
classes, although there is a significant number of tumors spatially distributed
between cluster groupings 7-9 and 2 on the UMAP projection that have pathway
characteristics of both. The top ten distinguishing pathways of class 2 tumors
were "cellular response to hypoxia," "regulation of mitotic cell cycle,"
"assembly of the pre-replicative complex," "metabolism of polyamines,"
"degradation of $\beta$-catenin by the destruction complex", "tumor necrosis
factor receptor 2 non-canonical _NF-$\kappa$B_ pathway", " _AUF1_ (hnRNP D0)
binds and destabilizes RNA", "regulation of apoptosis" and 3 other pathways
associated with aspects of hedgehog signaling. To determine once again whether
the genes driving pathway enrichment were commonly shared among these pathways
we plotted and performed unsupervised clustering on tumors with an individual
mutation matrix using every gene in each of the ten pathways, grouped by shared
pathway membership (**figure 3A**). Surprisingly, this class of tumors was not
predominated by 1 or more highly mutated genes across all cancers. Instead,
genes in the proteosomal pathway whose activity contributes to all the
aforementioned pathways and the majority of tumors in these classes have a
single mutation in one of these genes. We annotated the tumors in this class
according to tissue of origin to determine whether proteasome mutations have a
tissue-of-origin (or cancer-specific) bias, and indeed we could observe some
clustering. These data could explain how tissue-specific gene mutation
signatures might result in general cancer phenotypes by converging at the
pathway level.

The proteasome plays key roles in the transduction of signals from the p53, WNT
and NFkB pathways and it and proteasomal inhibitors have been approved for
clinical use in some patient populations [@proteasome_review_Jang]. To determine
whether this class of proteins is broadly sensitive to cancer-promoting
mutations, we tabulated each gene as prognostic or not in the human protein
expression atlas [@protein_atlas] and found 49 positive and 10 negative
associations ( _i.e._ higher expression associated with unfavorable outcomes is
"positive" _etc._) for 31 out of 43 (72%) proteasome genes (**figure 3B**).
Second we looked at survival in mutations (**figure 3C**). Only a handful of proteasome mutations were significantly associated with favorable or unfavorable outcomes, probably owing to lack of sufficient power.

Third, we checked the mutation classes using MAFtools [@maftools] and found the overwhelming
majority (x out of y) are missense _vs._ various other deleterious classes (see
**figure 3D**), suggesting that, similar to other oncogenic driver classes such
as KRAS and PIK3C, proteasomal subunit mutations may result in specific
dysregulation of the proteasomal complex activities that target cancer-promoting
pathways. Taken together, our data suggest that mutations in proteasomal
subunits are more widespread in cancer than previously appreciated and may
represent a targetable category of tumors.
</strike>

@@

As discussed previously, prior efforts to extract common signatures from pan-cancer 
datasets met with difficulty in distinguishing tumor samples from their tissue-specific 
-omics data signatures. Given our pathway-disruption based classification, this raises 
the question, are tumor phenotypes driven entirely by common driver genes, or by “silent”
tissue-specific effectors ( _i.e._ too few samples to detect above statistical significance 
thresholds), or a combination of both? To answer this question, we compared each cluster 
relative to the background to find deferentially mutated genes. We ranked the odd ratios
and selected the top ten enriched and depleted genes (pvalue < 0.01) for each cluster 
(**Figure 4**; odds ratios plot). Clusters 7 and 8, which shared multiple enrichment in 
signaling pathways, are largely driven by mutations in _PI3K_ and its homologs and _Ras_ 
genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary figure S5**). 
Interestingly, cluster 9, which also shared multiple enrichment in signaling pathways 
with clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. This result suggests 
that tumors in this cluster could be a derivative either cluster 7 or 8. Clusters 3 and 
5, defined by enrichment in metabolic pathways, had mutations in ribosomal proteins and 
nuclear pore complex, respectively. Cluster 4 had mutations in genes responsible for 
nucleosome structure. Cluster 2 had mutations in genes involved in protein degradation. 
Next, we investigated the proportion of samples per cancer type for the significant genes
within a cluster (**Figure 4**; heatmap). Surprisingly, cluster-specific tumors were not 
predominated by one or more highly mutated genes across all cancers. Instead, several 
cancer-specific tumors in these classes had mutation profiles that differ from one another 
( _e.g._ In cluster 2, LIHC was enriched for _PSME4_, KIRC was enriched for _UBC_, and 
PAAD was enriched for _PSMA1_). Taken together, our data identifies a framework cancer 
type specific mutation associated with a specific cluster.

@@

```{r child='survival-methods.Rmd'}
```
\newpage
```{r child='../text/fig_legend.Rmd'}
```

\newpage
```{r child='../code/figure1.Rmd'}
```
\newpage
```{r child='../code/figure2.Rmd'}
```
\newpage
```{r child = '../code/figure4.Rmd'}
```
\newpage
```{r child='../code/table2.Rmd'}
```
\newpage

![](../figures/fig6B.png)
\pagebreak
![](../figures/fig6C.png)

@@

```{r child='../code/supplements.Rmd'}
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, dpi=600, warning = FALSE)
suppressPackageStartupMessages(c(
  library(rstan),
  library(foreach),
  library(sqldf),
  library(tidyverse),
  library(ggthemes),
  library(data.table),
  library(biomaRt),
  library(dplyr),
  library(GenomicRanges),
  library(SummarizedExperiment),
  library(reactome.db),
  library(plyr),
  library(cluster),
  library(pheatmap),
  library(clusterProfiler),
  library(org.Hs.eg.db),
  library(dbscan),
  library(class),
  library(TCGAbiolinks),
  library(patchwork),
  library(ggplotify),
  library(ComplexHeatmap),
  library(knitr),
  library(kableExtra),
  library(UpSetR),
  library(cowplot),
  library(doRNG),
  library(uwot),
  library(janitor),
  library(FactoMineR),
  library(seriation),
  library(rtracklayer),
  library(poibin),
  library(flextable),
  library(plot3D)
))
```

@@

@@

As discussed previously, prior efforts to extract common signatures from pan-cancer 
datasets met with difficulty in distinguishing tumor samples from their tissue-specific 
-omics data signatures. Given our pathway-disruption based classification, this raises 
the question, are tumor phenotypes driven entirely by common driver genes, or by “silent”
tissue-specific effectors ( _i.e._ too few samples to detect above statistical significance 
thresholds), or a combination of both? To answer this question, we compared each cluster 
relative to the background to find deferentially mutated genes. We ranked the odd ratios
and selected the top ten enriched and depleted genes (pvalue < 0.01) for each cluster 
(**Figure 4**; odds ratios plot). Clusters 7 and 8, which shared multiple enrichment in 
signaling pathways, are largely driven by mutations in _PI3K_ and its homologs and _Ras_ 
genes, respectively (compare _PIK3CA_ and _KRAS_ panels of **Supplementary figure S5**). 
Interestingly, cluster 9, which also shared multiple enrichment in signaling pathways 
with clusters 7 and 8, is enriched for both _PIK3CA_ and _KRAS_. This result suggests 
that tumors in this cluster could be a derivative either cluster 7 or 8. Clusters 3 and 
5, defined by enrichment in metabolic pathways, had mutations in ribosomal proteins and 
nuclear pore complex, respectively. Cluster 4 had mutations in genes responsible for 
nucleosome structure. Cluster 2 had mutations in genes involved in protein degradation. 
We also observed genes that were enriched for a cluster are depleted from others ( _e.g._ 
_TP53_ is enriched in cluster 6, but depleted in cluster 7; _PIK3CA_ is enriched in cluster
7, but depleted in clusters 3 and 8). Next, we investigated the proportion of samples per 
cancer type for the significant genes within a cluster (**Figure 4**; heatmap). Surprisingly, 
cluster-specific tumors were not predominated by one or more highly mutated genes across all 
cancers. Instead, when observing the mutation rate for these genes within samples that belong
to a cluster, the mutation rate is heterogeneous across tumors by tissue origin ( _e.g._ In 
cluster 4, CESC was enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched
for _HIST1H2AC_). Even among the top most mutationally enriched genes within clusters there is
no global pattern, indicating that our classes are not driven by individual genes, but rather 
networks as a whole. Taken together, our data identifies a framework of cancer type-specific 
mutation associated with a specific cluster.

@@

```{r child='survival-methods.Rmd'}
```

```{r, results='asis'}
cat("\\newpage")
```

```{r child='../text/fig_legend.Rmd'}
```

```{r, results='asis'}
cat("\\newpage")
```
#### Figure 1
```{r child='../code/figure1v2.Rmd'}
```

```{r, results='asis'}
cat("\\newpage")
```
#### Figure 2
```{r child='../code/figure2v2.Rmd'}
```

```{r, results='asis'}
cat("\\newpage")
```
#### Figure 3
```{r child = '../code/figure3v2.Rmd'}
```

```{r, results='asis'}
cat("\\newpage")
```
#### Figure 4
```{r child = '../code/figure4v2.Rmd'}
```

```{r, results='asis'}
cat("\\newpage")
```
#### Figure 5
```{r child = '../code/figure2.Rmd'}
```

```{r, results='asis'}
cat("\\newpage")
```
#### Figure 6
```{r child = '../code/figure4.Rmd'}
```

```{r, results='asis'}
cat("\\newpage")
```
#### Table 1
```{r child = '../code/table2.Rmd'}
```

```{r, results='asis'}
cat("\\newpage")
```
![](../figures/fig6B.png)
\newpage
![](../figures/fig6C.png)

@@

```{r }
#child='../code/supplements.Rmd'
```
